---
title: "Text Analysis (2): Supervised Machine Learning"
subtitle: "J381M Week 5 Tutorial - Live Coding Version"
date: "February 12, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 8, fig.height = 5)
```

# Setup (Done Before Class)

```{r load-packages}
library(tidyverse)
library(tidytext)
library(textrecipes)
library(tidymodels)
```

---

# Part 3: Load and Prepare Data

## Load Dataset

```{r load-data}
# Load the Amazon Reviews CSV file
# File should be in the same folder as this RMD






```

## Clean Data

```{r clean-data}
# Clean the data:
# 1. Rename columns
# 2. Extract numeric rating
# 3. Filter valid data
# 4. Add word_count and sentiment columns










```

## Prepare Datasets

```{r prepare-datasets}
# Create two datasets:
# 1. Classification: Remove neutral, keep positive/negative
# 2. Regression: Keep all ratings

# CLASSIFICATION




# REGRESSION



```

---

# Part 4: CLASSIFICATION - Live Coding

## Visualize

```{r viz-class}
# Visualize class distribution






```

## Split Data

```{r split-class}
# Split 75-25 with stratification
# Hint: initial_split(data, prop = 0.75, strata = sentiment)






```

## Step 1: Preprocessing

```{r preprocess-class}
# Set up text preprocessing:
# - recipe(sentiment ~ review_text)
# - step_tokenize()
# - step_stopwords()
# - step_tokenfilter(max_tokens = 1000)
# - step_tfidf()








```

## Step 2: Model

```{r model-class}
# Choose logistic regression
# - logistic_reg()
# - set_engine("glm")
# - set_mode("classification")





```

## Step 3: Train

```{r train-class}
# Combine preprocessing + model
# - workflow()
# - add_recipe()
# - add_model()
# - fit()








```

## Step 4: Predict

```{r predict-class}
# Make predictions on test set
# - predict(new_data = test_class)
# - bind_cols(test_class)





```

## Step 5: Evaluate

```{r eval-class}
# Create confusion matrix
# - conf_mat(truth = sentiment, estimate = .pred_class)




# Calculate metrics
# - metrics(truth = sentiment, estimate = .pred_class)




# Compare to baseline



```

## Step 6: Feature Importance

```{r importance-class}
# Extract coefficients and visualize
# - extract_fit_parsnip()
# - tidy()
# - filter term != Intercept
# - slice_max(abs_estimate)









```

---

# Part 5: REGRESSION - Live Coding

## Visualize

```{r viz-reg}
# Visualize rating distribution






```

## Split Data

```{r split-reg}
# Split 75-25 for regression






```

## Step 1: Preprocessing

```{r preprocess-reg}
# Same preprocessing as classification
# But predict rating (numeric) instead of sentiment








```

## Step 2: Model

```{r model-reg}
# Choose linear regression
# - linear_reg()
# - set_engine("lm")
# - set_mode("regression")





```

## Step 3: Train

```{r train-reg}
# Combine preprocessing + model
# - workflow()
# - add_recipe()
# - add_model()
# - fit()








```

## Step 4: Predict

```{r predict-reg}
# Make predictions on test set





```

## Step 5: Evaluate

```{r eval-reg}
# Calculate metrics
# - metrics(truth = rating, estimate = .pred)




# Visualize predicted vs actual
# - ggplot with geom_jitter and geom_abline








```

---

*Tutorial by Sungwon Jung, J381M 2026*
