---
title: "Text Analysis (2): Supervised Machine Learning"
subtitle: "J381M Week 5 Tutorial - Instructor Version"
author: "Instructor: Sungwon Jung"
date: "February 12, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 8, fig.height = 5)
```

# Before We Begin: Setup 

## Install Required Packages

Before we start, we need to install some R packages. **Do this ONCE before class:**

```{r install-packages, eval=FALSE}
# Copy and paste this into your R console:
#install.packages(c(
#  "tidyverse",      # For data manipulation and visualization
#  "tidytext",       # For working with text data
#  "tidymodels",     # For machine learning
#  "textrecipes",
#  "stopwords"# For text preprocessing
#))

# Optional - for Exercise 3 later:
#install.packages("ranger")
```

**Having trouble?**
- **Mac users**: You might need to install Xcode tools if prompted
- **Windows users**: You might need Rtools
- **Everyone**: Try installing one package at a time if you get errors

---

## Load the Packages

Now let's load the packages we just installed:

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(textrecipes)
library(tidymodels)

cat("Packages loaded!\n")
```

## Quick Test

Let's make sure everything is working:

```{r test-setup}
# Create a simple test
test_data <- tibble(
  text = c("I love this", "This is bad", "Pretty good"),
  rating = c(5, 1, 4)
)

# Try to tokenize (split into words)
test_data %>% 
  unnest_tokens(word, text) %>%
  head()

cat("\nEverything works!\n")
```

If you see the table above with words, you're all set!

**Still having issues?** Ask for help now before we start!

---

# Part 1: Core ML Principles (10 min)

## What is Supervised Learning?

**Supervised Learning** = Learning from labeled examples

Think of it like learning with flashcards:
- **Front**: Input (review text)
- **Back**: Correct answer (rating or sentiment)
- **Goal**: Learn to predict the back from the front

## Two Main Tasks

### 1. Classification (Predicting Categories)

**Examples:**
- Is this review positive or negative?
- Is this email spam?
- Which topic does this article discuss?

**Output:** Discrete labels (positive, negative, spam, not spam)

### 2. Regression (Predicting Numbers)

**Examples:**
- What rating (1-5 stars) will this review give?
- How many likes will this post get?
- What's the temperature tomorrow?

**Output:** Continuous values (1.5 stars, 234 likes, 72°F)

## Train/Test Split

**Critical Rule:** Never test on data you trained on!

```
Total Data (100%)
│
├─ Training Set (75%)    ← Model learns here
└─ Test Set (25%)        ← Evaluate here (NEVER touch during training!)
```

**Why?** If you test on training data, you can't tell if your model actually learned patterns or just memorized answers!

### Dealing with Imbalanced Data

Real-world data is often **imbalanced** - one class appears much more than others.

**Example:** In our Amazon reviews:
- Negative reviews: 71% (14,350 reviews)
- Positive reviews: 29% (5,820 reviews)

**Problem:** A "dumb" model that always predicts "negative" gets 71% accuracy!

**Solutions:**
1. **Stratified sampling**: Keep same class proportions in train/test
   - We use `strata = sentiment` in `initial_split()`
2. **Look beyond accuracy**: Use precision, recall, F1-score
3. **Balanced metrics**: Consider performance on BOTH classes

**Today we'll use stratified sampling + balanced metrics!**

---

# Part 2: Text as Features

Computers don't understand words - we need to convert text to numbers.

## The Pipeline

```
"This product is amazing!" 
    ↓ Tokenization
["This", "product", "is", "amazing"]
    ↓ Remove Stop Words  
["product", "amazing"]
    ↓ Vectorization
[product: 1, amazing: 1, ...]
```

### Key Concepts

**1. Tokenization**: Split text into words

**2. Stop Words**: Common words with little meaning ("the", "a", "is")

**3. TF-IDF**: Weight words by importance
- **TF** (Term Frequency): How often word appears in document
- **IDF** (Inverse Document Frequency): How rare word is across all documents
- **TF-IDF**: Highlights unique, discriminative words

**Example:**
- "amazon" appears in every review → low TF-IDF (not useful)
- "excellent" appears only in positive reviews → high TF-IDF (very useful!)

---

# Part 3: Loading and Preparing Data

## Load Dataset

```{r load-data}
# Load Amazon Reviews
# Make sure the CSV file is in the same folder as this RMD file
# Or update the path below to where your file is located

amazon_reviews_raw <- read_csv(
  "/Users/sungwonjung/Downloads/Amazon_Reviews.csv",  # Change this path if needed
  col_types = cols(
    `Reviewer Name` = col_character(),
    Country = col_character(),
    Rating = col_character(),
    `Review Title` = col_character(),
    `Review Text` = col_character(),
    .default = col_character()
  )
)

cat("Total reviews:", nrow(amazon_reviews_raw), "\n")
glimpse(amazon_reviews_raw)
```

## Clean Data

```{r clean-data}
amazon_reviews_clean <- amazon_reviews_raw %>%
  rename(
    rating_text = Rating,
    review_text = `Review Text`
  ) %>%
  # Extract numeric rating
  mutate(rating = as.numeric(str_extract(rating_text, "\\d+"))) %>%
  # Filter valid data
  filter(
    !is.na(review_text),
    review_text != "",
    !is.na(rating),
    rating >= 1 & rating <= 5
  ) %>%
  # Add features
  mutate(
    word_count = str_count(review_text, "\\w+"),
    sentiment = case_when(
      rating >= 4 ~ "positive",
      rating <= 2 ~ "negative",
      TRUE ~ "neutral"
    )
  )

cat("After cleaning:", nrow(amazon_reviews_clean), "reviews\n")
```

## Prepare Datasets

```{r prepare-datasets}
# CLASSIFICATION: Binary sentiment (pos vs neg)
amazon_classification <- amazon_reviews_clean %>%
  filter(sentiment != "neutral") %>%
  mutate(sentiment = factor(sentiment, levels = c("negative", "positive"))) %>%
  select(review_text, sentiment, rating, word_count)

# REGRESSION: Predict rating (1-5)
amazon_regression <- amazon_reviews_clean %>%
  select(review_text, rating, sentiment, word_count)

cat("Classification dataset:", nrow(amazon_classification), "reviews\n")
cat("Regression dataset:", nrow(amazon_regression), "reviews\n")
```

## Example Reviews

```{r examples}
set.seed(2025)

cat("=== POSITIVE Examples ===\n\n")
amazon_classification %>%
  filter(sentiment == "positive") %>%
  slice_sample(n = 2) %>%
  pull(review_text) %>%
  str_trunc(200) %>%
  paste0(1:2, ". ", .) %>%
  cat(sep = "\n\n")

cat("\n\n=== NEGATIVE Examples ===\n\n")
amazon_classification %>%
  filter(sentiment == "negative") %>%
  slice_sample(n = 2) %>%
  pull(review_text) %>%
  str_trunc(200) %>%
  paste0(1:2, ". ", .) %>%
  cat(sep = "\n\n")
```

---

# Part 4: TASK 1 - CLASSIFICATION

Predicting sentiment (positive vs negative)

## Visualize the Data

```{r viz-class}
# Class distribution
ggplot(amazon_classification, aes(x = sentiment, fill = sentiment)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 5) +
  labs(title = "Classification: Class Distribution",
       x = NULL, y = "Count") +
  scale_fill_manual(values = c("negative" = "#F44336", "positive" = "#4CAF50")) +
  theme_minimal() +
  theme(legend.position = "none")

# Check class balance
class_dist <- amazon_classification %>% 
  count(sentiment) %>% 
  mutate(pct = round(n/sum(n)*100, 1))

cat("\n=== Class Distribution ===\n")
print(class_dist)

neg_count <- class_dist %>% filter(sentiment == "negative") %>% pull(n)
pos_count <- class_dist %>% filter(sentiment == "positive") %>% pull(n)
imbalance_ratio <- round(neg_count / pos_count, 1)

cat("\nCLASS IMBALANCE DETECTED!\n")
cat("Ratio:", imbalance_ratio, ": 1 (negative:positive)\n")
cat("Negative:", round(class_dist$pct[1], 1), "% | Positive:", round(class_dist$pct[2], 1), "%\n\n")

cat("Why this matters:\n")
cat("- A 'dumb' model that always predicts 'negative' would get", 
    round(class_dist$pct[1], 1), "% accuracy!\n")
cat("- Our model needs to do BETTER than this baseline\n")
cat("- We'll use stratified sampling to maintain balance in train/test splits\n\n")

# Feature analysis: Review length by sentiment
ggplot(amazon_classification, aes(x = sentiment, y = word_count, fill = sentiment)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Feature: Review Length by Sentiment",
       x = NULL, y = "Word Count") +
  scale_fill_manual(values = c("negative" = "#F44336", "positive" = "#4CAF50")) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Split Data

```{r split-class}
set.seed(2025)

# 75-25 split with stratification
split_class <- initial_split(amazon_classification, prop = 0.75, strata = sentiment)
train_class <- training(split_class)
test_class <- testing(split_class)

cat("Training:", nrow(train_class), "| Test:", nrow(test_class), "\n")
train_class %>% count(sentiment)
```

## Step 1: Set Up Text Preprocessing

```{r preprocess-class}
# Define how to convert text to numbers
preprocess_class <- recipe(sentiment ~ review_text, data = train_class) %>%
  step_tokenize(review_text) %>%              # Split into words
  step_stopwords(review_text) %>%             # Remove "the", "a", etc.
  step_tokenfilter(review_text, max_tokens = 1000) %>%  # Keep top 1000 words
  step_tfidf(review_text)                     # Calculate word importance

preprocess_class
```

## Step 2: Choose Model

```{r model-class}
# Use Logistic Regression
model_class <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```

## Step 3: Train the Model

```{r train-class}
# Combine preprocessing + model
pipeline_class <- workflow() %>%
  add_recipe(preprocess_class) %>%
  add_model(model_class)

# Train
fit_class <- pipeline_class %>%
  fit(data = train_class)

cat("Classification model trained!\n")
```

## Step 4: Make Predictions

```{r predict-class}
# Predict on test set
pred_class <- fit_class %>%
  predict(new_data = test_class) %>%
  bind_cols(test_class)

# View some predictions
pred_class %>%
  select(review_text, sentiment, .pred_class) %>%
  slice_sample(n = 10)
```

## Step 5: Evaluate Performance

```{r eval-class}
# Confusion Matrix
conf_mat_class <- pred_class %>%
  conf_mat(truth = sentiment, estimate = .pred_class)

print(conf_mat_class)

autoplot(conf_mat_class, type = "heatmap") +
  labs(title = "Confusion Matrix - Sentiment Classification") +
  theme_minimal()

# Basic metrics
metrics_class <- pred_class %>%
  metrics(truth = sentiment, estimate = .pred_class)

print(metrics_class)

accuracy <- metrics_class %>% filter(.metric == "accuracy") %>% pull(.estimate)
cat("\nOverall Accuracy:", round(accuracy * 100, 2), "%\n")

# Compare to baseline
baseline_accuracy <- neg_count / (neg_count + pos_count)
cat("\n=== Baseline Comparison ===\n")
cat("Baseline (always predict negative):", round(baseline_accuracy * 100, 2), "%\n")
cat("Our model:", round(accuracy * 100, 2), "%\n")
cat("Improvement:", round((accuracy - baseline_accuracy) * 100, 2), "percentage points\n")

if (accuracy > baseline_accuracy + 0.10) {
  cat("\nModel is significantly better than baseline!\n")
} else {
  cat("\nModel barely beats the baseline - might need improvement\n")
}

# Class-specific metrics (important for imbalanced data!)
cat("\n=== Class-Specific Performance ===\n")
class_metrics <- pred_class %>%
  precision(truth = sentiment, estimate = .pred_class) %>%
  bind_rows(pred_class %>% recall(truth = sentiment, estimate = .pred_class)) %>%
  bind_rows(pred_class %>% f_meas(truth = sentiment, estimate = .pred_class))

print(class_metrics)

cat("\nWhy these metrics matter with imbalanced data:\n")
cat("- Precision: Of predictions we made, how many were correct?\n")
cat("- Recall: Of actual cases, how many did we catch?\n")
cat("- F1: Harmonic mean of precision and recall\n")
```

## Step 6: Feature Importance

```{r importance-class}
# Extract coefficients
coefs <- fit_class %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = str_remove(term, "tfidf_review_text_"),
    abs_estimate = abs(estimate)
  )

# Top 20 words
top20 <- coefs %>% slice_max(abs_estimate, n = 20)

ggplot(top20, aes(x = reorder(term, estimate), y = estimate, fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  labs(title = "Most Important Words for Classification",
       subtitle = "Green = Positive indicators | Red = Negative indicators",
       x = "Word", y = "Coefficient") +
  scale_fill_manual(values = c("TRUE" = "#4CAF50", "FALSE" = "#F44336")) +
  theme_minimal() +
  theme(legend.position = "none")

cat("\n=== Top Positive Words ===\n")
coefs %>% slice_max(estimate, n = 5) %>% select(term, estimate) %>% print()

cat("\n=== Top Negative Words ===\n")
coefs %>% slice_min(estimate, n = 5) %>% select(term, estimate) %>% print()
```

---

# Part 5: TASK 2 - REGRESSION

Predicting rating (1-5 stars)

## Visualize the Data

```{r viz-reg}
# Target distribution
ggplot(amazon_regression, aes(x = factor(rating))) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 5) +
  labs(title = "Regression: Target Distribution",
       subtitle = "Predicting rating (1-5 stars)",
       x = "Rating", y = "Count") +
  theme_minimal()

# Feature analysis: Review length by rating
ggplot(amazon_regression, aes(x = factor(rating), y = word_count, fill = factor(rating))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Feature: Review Length by Rating",
       x = "Rating", y = "Word Count") +
  scale_fill_brewer(palette = "RdYlGn", direction = 1) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Split Data

```{r split-reg}
set.seed(2025)

split_reg <- initial_split(amazon_regression, prop = 0.75)
train_reg <- training(split_reg)
test_reg <- testing(split_reg)

cat("Training:", nrow(train_reg), "| Test:", nrow(test_reg), "\n")
```

## Step 1: Set Up Text Preprocessing

```{r preprocess-reg}
# Same preprocessing as classification
preprocess_reg <- recipe(rating ~ review_text, data = train_reg) %>%
  step_tokenize(review_text) %>%
  step_stopwords(review_text) %>%
  step_tokenfilter(review_text, max_tokens = 1000) %>%
  step_tfidf(review_text)
```

## Step 2: Choose Model

```{r model-reg}
# Use Linear Regression
model_reg <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")
```

## Step 3: Train the Model

```{r train-reg}
# Combine preprocessing + model
pipeline_reg <- workflow() %>%
  add_recipe(preprocess_reg) %>%
  add_model(model_reg)

# Train
fit_reg <- pipeline_reg %>%
  fit(data = train_reg)

cat("Regression model trained!\n")
```

## Step 4: Make Predictions

```{r predict-reg}
# Predict
pred_reg <- fit_reg %>%
  predict(new_data = test_reg) %>%
  bind_cols(test_reg) %>%
  mutate(error = rating - .pred)

# View some predictions
pred_reg %>%
  select(rating, .pred, review_text) %>%
  mutate(review_text = str_trunc(review_text, 80)) %>%
  slice_sample(n = 10)
```

## Step 5: Evaluate Performance

```{r eval-reg}
# Metrics
metrics_reg <- pred_reg %>%
  metrics(truth = rating, estimate = .pred)

print(metrics_reg)

rmse <- metrics_reg %>% filter(.metric == "rmse") %>% pull(.estimate)
rsq <- metrics_reg %>% filter(.metric == "rsq") %>% pull(.estimate)

cat("\nRMSE:", round(rmse, 3), "stars\n")
cat("R²:", round(rsq, 3), "\n")

# Predicted vs Actual
ggplot(pred_reg, aes(x = rating, y = .pred)) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Regression: Predicted vs Actual",
       subtitle = paste0("RMSE = ", round(rmse, 2), " | R² = ", round(rsq, 2)),
       x = "Actual Rating", y = "Predicted Rating") +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(breaks = 1:5) +
  theme_minimal()

# Error distribution
ggplot(pred_reg, aes(x = error)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Prediction Errors",
       x = "Error (Actual - Predicted)", y = "Count") +
  theme_minimal()

# Errors by rating
ggplot(pred_reg, aes(x = factor(rating), y = abs(error))) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Which Ratings Are Hardest to Predict?",
       x = "Actual Rating", y = "Absolute Error") +
  theme_minimal()
```

## Step 6: Feature Importance

```{r importance-reg}
coefs_reg <- fit_reg %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = str_remove(term, "tfidf_review_text_"),
    abs_estimate = abs(estimate)
  )

top20_reg <- coefs_reg %>% slice_max(abs_estimate, n = 20)

ggplot(top20_reg, aes(x = reorder(term, estimate), y = estimate, fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  labs(title = "Most Important Words for Rating Prediction",
       subtitle = "Positive = Higher ratings | Negative = Lower ratings",
       x = "Word", y = "Coefficient") +
  scale_fill_manual(values = c("FALSE" = "#F44336", "TRUE" = "#4CAF50")) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Part 6: Comparing Classification vs Regression

```{r compare}
cat("============================================\n")
cat("CLASSIFICATION vs REGRESSION\n")
cat("============================================\n\n")

cat("CLASSIFICATION (Binary: Positive vs Negative)\n")
cat("- Task: Predict category\n")
cat("- Accuracy:", round(accuracy * 100, 2), "%\n")
cat("- Use when: Clear binary/multi-class decisions\n")
cat("- Examples: Spam filter, sentiment analysis, topic classification\n\n")

cat("REGRESSION (Continuous: 1-5 stars)\n")
cat("- Task: Predict numeric value\n")
cat("- RMSE:", round(rmse, 3), "stars | R²:", round(rsq, 3), "\n")
cat("- Use when: Predicting quantities or scores\n")
cat("- Examples: Price prediction, rating prediction, demand forecasting\n\n")

cat("Key Differences:\n")
cat("1. Output: Category vs Number\n")
cat("2. Metrics: Accuracy/F1 vs RMSE/R²\n")
cat("3. Model: Logistic Regression vs Linear Regression\n")
cat("4. Interpretation: Probability of class vs Expected value\n")
```

---

# Wrap-up

## Key Takeaways

- **Supervised Learning**: Learn from labeled examples
- **Classification**: Predict categories (positive/negative)
- **Regression**: Predict numbers (1-5 stars)
- **Text Preprocessing**: Tokenize → Remove stop words → TF-IDF
- **Evaluation**: Accuracy for classification, RMSE for regression
- **Feature Importance**: Which words matter most?
- **Class Imbalance**: Real data is often imbalanced - use stratified sampling and balanced metrics

## Next Steps

- Try with your own text data
- Experiment with different models
- Learn about cross-validation
- Explore deep learning for NLP

## Resources

- [Tidymodels](https://www.tidymodels.org/)
- [Text Mining with R](https://www.tidytextmining.com/)
- [Supervised ML for Text](https://smltar.com/)

---

*Tutorial by Sungwon Jung, J381M 2026*
