---
title: "J381M_H01_R_basics"
date: "2026-01-13"
output: 
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, gridExtra, ggrepel, plotly, skimr) 
# install a package if it does not exist already and put the package in the path (library)
# dplyr, ggplot2,tidyr
```

\pagebreak

# Objectives {-}

Data Science is a field of science connecting statistics, computer science, and domain knowledge. We would like to discover the pattern of differences and changes, as well as the reasons behind the scene.

For any well-designed study, we need:

1. a clear goal of the study

Using domain knowledge we may list possible factors related to the study, i.e., we need to first design what information may help us to achieve the goal. Taking feasibility and cost into account, we will either collect original data of interest (e.g., from experiments, surveys, or interviews), or we may learn important insights from existing data (e.g., from public databases, or from other studies). In either case, we need to ensure the quantity and quality of data, which determines the success of the study. 

Once we have the data, we proceed to extract useful information. To use the data correctly and efficiently we must understand the data first. In this lecture, we go through some basic data acquisition/preparation and exploratory data analysis to understand the nature of the data, and to explore plausible relationships among the variables. We defer the formal modeling later. 




Data mining tools have been expanding dramatically in the past 20 years. We inevitably need to use computing software. R is one of the most popular software among data scientists and in academia. It is an open-source programming language so users can customize existing codes or functions according to their needs. Most state-of-the-art methodologies are implemented as R packages. 

While the number of scientific papers is soaring, especially during COVID, a significant amount of the studies can not be reproduced or replicated. This phenomenon is an ongoing crisis termed the replicability crisis. In an effort to produce trustworthy and reproducible results, we use R-markdown. It achieves many goals: 1) Anyone can rerun our study to replicate the results 2) We will be able to run our data analysis and produce reports at the same time. Communication between us and readers/decision-makers is essential. 

We focus on data preparation, data cleaning, and exploratory data analyses (EDA) in this module.

R combined with R-markdown will be used. Please go through advanced_R_tutorial.Rmd to learn a set of extremely useful EDA tools such as `dplyr`, `ggplot` and `data.table` and more.

This lecture is rich and rather involved with both data analyses and R-packages/functions. Perhaps you may read through a compiled file first to grasp the main theme then come back to this .rmd file to run line by line. That is one way you will turn the coding to yourself!


Contents:

0. Suggested extra readings/doing:  
    + run and study Get_staRted.Rmd
    + run and study advanced_R_tutorial.Rmd and advanced_R_tutorial.html
    + read `50 years of data science` and `Teaching data science` available in canvas. 
    + Data set: `MLPayData_Total.csv`
1. Case Study: Billion dollar Billy Beane
2. Study flow:
    + Study design
    + Gathering data
    + Process data (tidy data)
    + Exploratory Data Analysis (EDA)
    + Conclusion/Challenges
4. R functions 
    + basic r functions
    + dplyr
    + ggplot

**Handy cheat sheets**

[DPLYR Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

[ggplot Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)










```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
